---
title: "Sleep stage composition & brain health"
format:
  html:
    theme: cosmo
    css: report.css
    embed-resources: true
    toc: true
    toc-title: "In this report"
    toc-depth: 2
    title-block-banner: "#173f4f"
    title-block-banner-color: "#ffffff"
    smooth-scroll: true
    page-layout: full
    grid:
      body-width: 1400px
      sidebar-width: 220px
execute:
  echo: false
  message: false
  warning: false
  cache: false
params:
  targets_store: "_targets"
---

```{r}
#| label: setup
#| include: false
library(plotly)
library(gtsummary)
```

This report describes the eligible analysis cohort before imputation and the
distribution of sleep stages at SHHS-2.

## Sleep-stage composition

The five panels show quintiles of REM as a proportion of total sleep time.
N1, N2, N3, and REM sum to 100%; WASO is excluded from this descriptive view.
The analysis itself uses a five-part composition that includes WASO.

::: {.callout-tip appearance="simple"}
Drag within a panel to rotate it, scroll to zoom, and hover over a marker to see
its composition-bin ranges and participant count. Marker area represents count;
the coloured triangles mark the REM-quintile boundaries.
:::

```{r}
#| label: composition-distribution
#| out-width: "100%"
targets::tar_read(comp_dist_plot, store = params$targets_store)
```

The display aggregates compositions into two-percentage-point bins, shown at
their lower corners. Tied REM proportions remain together, so panel sizes can
differ. The four components have three independent dimensions: REM is the
remainder after N1, N2, and N3.

## Cohort characteristics

The table describes the model covariates, presenting sleep stages and WASO in raw
minutes at SHHS-1 and SHHS-2 in place of the ILR coordinates. It also includes the
three follow-up outcomes: cognitive summary score, hippocampal volume, and cerebral
brain volume. Continuous
variables are summarised as mean (SD), and categorical variables as counts and
percentages among non-missing observations. Missing counts are shown separately.
The baseline and follow-up cognitive scores are derived from recorded assessments before imputation.
Categorical codes are retained where descriptive labels have not been established.

```{r}
#| label: cohort-covariates
targets::tar_read(covariate_table, store = params$targets_store) |>
  gtsummary::as_gt()
```
