---
title: "Sleep stage composition & brain health"
format:
  html:
    theme: cosmo
    css: report.css
    embed-resources: true
    toc: true
    toc-title: "In this report"
    toc-depth: 2
    title-block-banner: "#173f4f"
    title-block-banner-color: "#ffffff"
    smooth-scroll: true
    page-layout: full
    grid:
      body-width: 1400px
      sidebar-width: 220px
execute:
  echo: false
  message: false
  warning: false
  cache: false
params:
  targets_store: "_targets"
---

```{r}
#| label: setup
#| include: false
library(plotly)
library(gtsummary)
```

This report describes the eligible analysis cohort before imputation and the
distribution of sleep stages at SHHS-2.

## Sleep-stage composition

The five panels show quintiles of REM as a proportion of total sleep time.
N1, N2, N3, and REM sum to 100%. The analysis uses these four sleep stages
as the composition and adjusts for SHHS-2 WASO separately.

::: {.callout-tip appearance="simple"}
Drag within a panel to rotate it, scroll to zoom, and hover over a marker to see
its composition-bin ranges and participant count. Marker area represents count;
the coloured triangles mark the REM-quintile boundaries.
:::

```{r}
#| label: composition-distribution
#| out-width: "100%"
targets::tar_read(comp_dist_plot, store = params$targets_store)
```

The display aggregates compositions into two-percentage-point bins, shown at
their lower corners. Tied REM proportions remain together, so panel sizes can
differ. The four components have three independent dimensions: REM is the
remainder after N1, N2, and N3.

## Cohort characteristics

The table describes the model covariates, presenting sleep stages and WASO in raw
minutes at SHHS-1 and SHHS-2 in place of the ILR coordinates. It also includes the
three follow-up outcomes: cognitive summary score, hippocampal volume, and cerebral
brain volume. Continuous
variables are summarised as mean (SD), and categorical variables as counts and
percentages among non-missing observations. Missing counts are shown separately.
The baseline and follow-up cognitive scores are derived from recorded assessments before imputation.
Categorical codes are retained where descriptive labels have not been established.

```{r}
#| label: cohort-covariates
targets::tar_read(covariate_table, store = params$targets_store) |>
  gtsummary::as_gt()
```

## Sleep window reallocations

Each plot shows the estimated mean difference as minutes shift between N3 and
another sleep stage. Positive minutes shift time toward N3; negative minutes
shift time away from N3. The shaded bands are 95% confidence intervals, and
the dotted line marks no difference.

```{r}
#| label: setup-pair-plots
#| include: false
pair_plots <- targets::tar_read(
  estimates_summary_pair_plot,
  store = params$targets_store
)
show_pair_plots <- function(outcome) {
  plots <- Filter(\(plot) plot$data$outcome[[1]] == outcome, pair_plots)
  for (plot in plots) print(plot + ggplot2::labs(title = NULL))
}
```

::: {.panel-tabset}

### Cognitive summary score

```{r}
#| label: pair-plots-cognition
#| layout-ncol: 2
#| fig-width: 7
#| fig-height: 5.5
#| fig-cap: "Estimated mean differences in cognitive summary score."
#| fig-subcap: ["N1 and N3", "N2 and N3", "REM and N3"]
show_pair_plots("pc1_s2")
```

### Hippocampal volume

```{r}
#| label: pair-plots-hippocampus
#| layout-ncol: 2
#| fig-width: 7
#| fig-height: 5.5
#| fig-cap: "Estimated mean differences in hippocampal volume."
#| fig-subcap: ["N1 and N3", "N2 and N3", "REM and N3"]
show_pair_plots("Hippo_s2")
```

### Cerebral brain volume

```{r}
#| label: pair-plots-cerebrum
#| layout-ncol: 2
#| fig-width: 7
#| fig-height: 5.5
#| fig-cap: "Estimated mean differences in cerebral brain volume."
#| fig-subcap: ["N1 and N3", "N2 and N3", "REM and N3"]
show_pair_plots("Cerebrum_tcb_s2")
```

:::

## Best and worst compositions

Training and test sets each contain half the participants. Within training
data, the search selects the best and worst four-stage compositions, both
without a TST restriction and at specified TST values. The selected fixed
compositions are then evaluated in test data. SHHS-2 WASO remains at each
participant's observed value throughout.

## Best and worst compositions by total sleep time

Each table shows the best and worst composition selected in each training split
at the specified total sleep time. All durations are in minutes.

```{r}
#| label: setup-extreme-compositions
#| include: false
extreme_compositions <- as.data.frame(targets::tar_read(
  ideal_split_extremes_by_tst,
  store = params$targets_store
))
show_extreme_compositions <- function(outcome_name) {
  rows <- extreme_compositions[extreme_compositions$outcome == outcome_name, ]
  rows <- rows[
    order(match(rows$policy, c("best", "worst")), rows$tst_minutes, rows$split_id),
    c("policy", "tst_minutes", "n1_s2", "n2_s2", "n3_s2", "rem_s2")
  ]
  names(rows) <- c("policy", "tst_minutes", "n1", "n2", "n3", "rem")
  knitr::kable(rows, row.names = FALSE)
}
```

::: {.panel-tabset}

### Cognitive summary score

```{r}
#| label: extreme-compositions-cognition
show_extreme_compositions("pc1_s2")
```

### Hippocampal volume

```{r}
#| label: extreme-compositions-hippocampus
show_extreme_compositions("Hippo_s2")
```

### Cerebral brain volume

```{r}
#| label: extreme-compositions-cerebrum
show_extreme_compositions("Cerebrum_tcb_s2")
```

:::
